Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-18 paper/paper.tex

Indeed, there were reports of cases that far preceded the outbreak at the Huanan Seafood Market.
The \textit{Lancet} described a confirmed case having no association with the market whose symptoms began on December 1, 2019~\citep{huang2020clinical}.
The \textit{South China Morning Post} described nine cases from November 2019 including details on patient age and sex, noting that none were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
After this, reports of early cases slowed---and in 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.

@@

The main \texttt{Snakemake} pipeline then downloads relevant SRA files from the Google Cloud links by using \texttt{wget} to download files from the Google Cloud from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs beyond the scope of this study indicates that only sometimes are deleted SRA files still available via the Google Cloud.
For this reason, key runs have been archived on the Wayback Machine as described above, and all downloaded SRA files relevant to this study are included in the GitHub repository.
Note also that as described in this paper's main text, two SRA files could not be downloaded from the Google Cloud using the aforementioned method, and so are not part of this study.

@@

I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to generate files covering the coverage at each site by aligned nucleotides with a quality score of at least 20.
These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the coverage plot that enable zooming and mouseovers to get statistics for specific sites.
I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} was designed to only cover part of the genome.

I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.

@@

\subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
I manually

>>>>

Indeed, there were reports of cases that far preceded the outbreak at the Huanan Seafood Market.
The \textit{Lancet} described a confirmed case having no association with the market whose symptoms began on December 1, 2019~\citep{huang2020clinical}.
The \textit{South China Morning Post} described nine cases from November 2019 including details on patient age and sex, noting that none were confirmed to be ``patient zero''~\citep{ma2020SCMP}.
Professor Yu Chuanhua of Wuhan University told the \textit{Health Times} that records he reviewed showed two cases in mid-November, and one suspected case on September 29~\citep{healthtimes2020}.
At about the same time as Professor Chuanhua's interview, the Chinese CDC issued an order forbidding sharing of information about the COVID-19 epidemic without approval~\citep{chinacdc2020}, and shortly thereafter Professor Chuanhua re-contacted the \textit{Health Times} to say the November cases could not be confirmed~\citep{healthtimes2020}.
Then China's State Council issued a much broader order requiring central approval of all publications related to COVID-19 to ensure they were coordinated ``like moves in a game of chess''~\citep{Kang2020}.
In 2021, the joint WHO-China report dismissed all reported cases prior to December 8 as not COVID-19, and revived the theory that the virus might have originated at the Huanan Seafood Market~\citep{WHO2021origins}.

@@

I recovered the SRA files from the Google Cloud by using \texttt{wget} to download files with from paths like \path{https://storage.googleapis.com/nih-sequence-read-archive/run/SRR11313485/SRR11313485}.
Note that I cannot guarantee that these Google Cloud links will remain active, as my analyses of other deleted SRA runs (beyond the scope of this study) indicates that only sometimes are deleted SRA files still available via the Google Cloud.
For this reason, key runs have been archived on the Wayback Machine as described above, and all downloaded SRA files relevant to this study are included in the GitHub repository.
Note also that as described in this paper's main text, two SRA files could not be downloaded from the Google Cloud using the aforementioned method, and so are not part of this study.

@@

I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to determine the coverage at each site by aligned nucleotides with a quality score of at least 20.
These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the plots that enable zooming and mouseovers to get statistics for specific sites.
I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} only covers part of the genome.

I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
These sequences in the region of interest (21,570 to 29,550) are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/results/recovered_seqs.fa}.

@@

\subsection{Curation and analysis of early SARS-CoV-2 sequences from GISAID}
For the broader analyses of existing SARS-CoV-2 sequences, I downloaded all sequences from collected prior to March of 2020 from GISAID.
The accessions of these sequences are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/accessions.txt}, and a table acknowledging the labs and authors is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/gisaid_sequences_through_Feb2020/acknowledgments.csv}.

I then used \texttt{mafft}~\citep{katoh2013mafft} to align these sequences to the proCoV2 reference described above, stripped any sites that were gapped relative to the reference, and filtered the sequences using the following criteria:
\begin{itemize}
\item I removed any sequences collected after February 28, 2020.
\item I removed any sequences that had $\ge$4 mutations within any 10-nucleotide stretch, as such runs of mutations often indicate sequencing errors.
\item I removed any sequence for which the alignment covered $<$90\% of the proCoV2 sequence.
\item I removed any sequence with $\ge$15 mutations relative to the reference.
\item I removed any sequence with $\ge$5,000 ambiguous nucleotides.
\end{itemize}

I then annotated the sequences using some additional information.
First, I annotated sequences based on the joint WHO-China report~\citep{WHO2021origins} and also \citet{zhu2020brief} to keep only one representative from multiply sequenced patients, and to indicate which sequences were from patients associated with the Huanan Seafood Market.
My version of these annotations is at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/WHO_China_Report_Dec2019_cases.yaml}.
Next, I identified some sequences in the set that were clearly duplicates from the same patient, and removed these.
The annotations used to remove these duplicates are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/seqs_to_exclude.yaml}.
Finally, I used information from ~\citet{chan2020familial} and \citet{kang2020evidence} to identify patients who were infected in Wuhan before January 5 of 2020, but ultimately sequenced in Guangdong: these annotations are at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/blob/main/data/Wuhan_exports.yaml}.

I next removed any of the handful of mutations noted by \citet{turakhia2020stability} to be lab artifacts that commonly afflict SARS-CoV-2 sequences.
I also limited the analyses to the region of the genome that spans from the start of the first coding region (ORF1ab) to the end of the last (ORF10), because I noticed that some sequences had suspicious patterns (such as many mutations or runs of mutations) near the termini of the genome.

The plot in Figure~\ref{fig:deltadist_RaTG13} contains all of the GISAID sequences after this filtering, and was rendered using \texttt{Altair}~\citep{vanderplas2018altair}.
The plot in Figure~\ref{fig:deltadist_jitter} shows the filtered GISAID sequences collected before February of 2020 plus the 13 good coverage recovered partial early outpatient sequences (Table~\ref{tab:mutations}), considering only the region covered by the partial sequences (21,570 to 29,550).

\subsection{Phylogenetic analyses}
